---
title: "Wilson and Obdeyn (2009)"
output: html_document
date: "2026-03-04"
---

# Wilson and Obdeyn's schematic example

The purpose of this example is to show that if you generate observed counts using marginal probabilities and a cooccurrence restriction, calculating O/E does not recover the true probability.

First we construct the table of generating probabilities

```{r}
# Construct the table of generating probabilities
segs = c("p","t","k")
freqs = c(1/3,1/2,1/6)
wilsonobdeyn2008 = data.frame(cbind(expand.grid(segs,segs,stringsAsFactors=TRUE),
expand.grid(freqs,freqs)))
names(wilsonobdeyn2008) = c("C1","C2","C1.prob","C2.prob")
wilsonobdeyn2008$CC.prob = c(.5,1,1,1,.5,1,1,1,.5)
wilsonobdeyn2008$gen.prob = with(wilsonobdeyn2008, C1.prob*C2.prob*CC.prob)

```

Now normalize to get the probability of each cell

```{r}
wilsonobdeyn2008$gen.prob = wilsonobdeyn2008$gen.prob/sum(wilsonobdeyn2008$gen.prob)
```

Calculate the expected counts, for a sample of 5000 words

```{r}
# Expected counts (frequencies) in sample of 5000
wilsonobdeyn2008$Freq = round(wilsonobdeyn2008$gen.prob*5000)

```

Now we are ready to see the O/E problem that Wilson and Obdeyn point out.

Using the table of expected counts that we just generated, we calculate the actual probabilities of each segment

```{r}
wilsonobdeyn2008.obsfreq = matrix(wilsonobdeyn2008$Freq,nrow=3)
wilsonobdeyn2008.obsprob.C1 = data.frame(prob=rowSums(wilsonobdeyn2008.obsfreq)/sum(wilsonobdeyn2008.obsfreq),row.names=c("p","t","k"))
wilsonobdeyn2008.obsprob.C2 = data.frame(prob=colSums(wilsonobdeyn2008.obsfreq)/sum(wilsonobdeyn2008.obsfreq),row.names=c("p","t","k"))

wilsonobdeyn2008$C1.obsprob = wilsonobdeyn2008.obsprob.C1[wilsonobdeyn2008$C1,"prob"]
wilsonobdeyn2008$C2.obsprob = wilsonobdeyn2008.obsprob.C2[wilsonobdeyn2008$C2,"prob"]

```

Now calculate O/E. Not only does this not yield 1's and .5's, but it actually yields different degrees of underattested by place (.58, .72, .48), making it look like a subsidiary feature is active.

```{r}
wilsonobdeyn2008$OE = wilsonobdeyn2008$Freq / (wilsonobdeyn2008$C1.obsprob*wilsonobdeyn2008$C2.obsprob*5000)
round(wilsonobdeyn2008$OE,2)

```

# Using loglinear models instead

We create columns for factors to test for different degrees of restriction. The "baseline" is a general OCP place effect, but we are interested in seeing whether we get a better fit by allowing the model to specifically differentiate OCP labial or OCP dorsal.

```{r}
wilsonobdeyn2008$OCP_Place = c(1,0,0,0,1,0,0,0,1)
wilsonobdeyn2008$OCP_Lab = c(1,0,0,0,0,0,0,0,0)
wilsonobdeyn2008$OCP_Cor = c(0,0,0,0,1,0,0,0,0)
wilsonobdeyn2008$OCP_Dors = c(0,0,0,0,0,0,0,0,1)

```

Now we set up the contrasts so we can model the counts with a loglinear (poisson) model.

```{r}
str(wilsonobdeyn2008$C1)
contrasts(wilsonobdeyn2008$C1)

```

A simple model, just to show the syntax/shape of what we want to construct

```{r}
wilsonobdeyn2008.simplemodel = glm(Freq ~ C1+C2, family=poisson(link="log"),data=wilsonobdeyn2008)
summary(wilsonobdeyn2008.simplemodel)

```

Now a model that incorporates just OCP place, to compare with one that also includes specific OCP constraints. For 3 places, we can test for 2 effects; here, I've arbitrarilychosen Lab and Cor

```{r}
wilsonobdeyn2008.genonly = glm(Freq ~ C1 + C2 + OCP_Place, data=wilsonobdeyn2008,family=poisson(link="log"))
wilsonobdeyn2008.genandspec = glm(Freq ~ C1 + C2 + OCP_Place + OCP_Lab + OCP_Cor, data=wilsonobdeyn2008,family=poisson())

anova(wilsonobdeyn2008.genonly,wilsonobdeyn2008.genandspec,test="Chisq")

```

Interpreting the coefficients for the C1 frequencies.

```{r}
# This is maybe easier with dummy coding
contrasts(wilsonobdeyn2008$C1) = contr.treatment(levels(wilsonobdeyn2008$C1))
contrasts(wilsonobdeyn2008$C2) = contr.treatment(levels(wilsonobdeyn2008$C2))

wilsonobdeyn2008.OCPgen = glm(Freq ~ C1 + C2 + OCP_Place, data=wilsonobdeyn2008,family=poisson())
summary(wilsonobdeyn2008.OCPgen)

# t vs. p: 1.5 times more frequent
exp(0.40515)
# k vs. p: 1/2 as frequent
exp(-0.69335)

wilsonobdeyn2008.OCPspec = glm(Freq ~ C1 + C2 + OCP_Place + OCP_Lab + OCP_Cor, data=wilsonobdeyn2008,family=poisson())
summary(wilsonobdeyn2008.OCPspec)

anova(wilsonobdeyn2008.OCPgen,wilsonobdeyn2008.OCPspec,test="Chisq")


```
