---
title: "BCM"
format: html
editor: visual
---

```{r}
library(tidyverse)
```

## Fitting Biased Choice Models as log-linear models

Read in the Redford & Diehl confusion matrix data

```{r}
rd = read_csv("redford-diehl-exp1.csv")

# Make C#V the baseline context:
rd$context = factor(rd$context, levels=c("C#V","CV","C#C"))
# Declare stimulus and response as factors
rd$stimulus = factor(rd$stimulus)
rd$response = factor(rd$response)
```

For initial analyses, we will only look at the data involving stops /p, t, k/

```{r}
stops = c("p","t","k")
rdstops = filter(rdfull, stimulus %in% stops & response %in% stops)
rdstops

# reset levels of stimulus and response
rdstops$response = factor(rdstops$response)
rdstops$stimulus = factor(rdstops$stimulus)
```

## Generating distance dummy variables

To fit a BCM, we need to generate dummy variable to encode the relationships required to fit distances - i.e. dummy variables for each pair of non-identical stimulus categories that have value 1 where stimulus and response involve that pair of categories (in either order).

The following function automatically generates a matrix, d, of dummy variables for a data set. The arguments are the dataframe and the names of the variable encoding stimulus category and the name of the variable encoding response category (by default, "stimulus" and "response")

```{r}
distmatrix=function(data, stimulus_var="stimulus", response_var="response"){
# 'cats' is the list of categories
cats = levels(data[[stimulus_var]])
n = length(cats)
# initialize distance matrix, d, as an array of 0s, with the same number of rows as data, 
# and enough columns for the dummy distance variables
d = array(0,dim=c(length(data[[stimulus_var]]),0.5*n*(n-1)))
# initialize a vector to hold the names of the dummy distance variables
dvars = rep.int(0,0.5*n*(n-1))
# initialize col, to keep track of which column of d is currently being populated
col=0
# Main loop: Enumerate all the pairs of distinct stimulus categories, and define a column of d for each
# loop through the cats list from first to penultimate item on the list.
for (i in 1:(n-1)){
  # cat1 is the ith category
  cat1=cats[i]
  # for each cat1, loop through the cats list from the next category to the end of the list
  for (j in (i+1):n){
    # increment col, the current column of d
    col=col+1
    # cat2 is the jth category
    cat2=cats[j]
    # the name of the current column is the concatenation of cat1 and cat2 joined by a period
    dvars[col]=paste(cat1,cat2,sep=".")
    # for each row in the dataframe, the cat1cat2 distance variable has value 1 if stimulus is cat1 and response is cat2
    # or stimulus is cat2 and response is cat1. Otherwise the value remains at 0.
    for (i in 1:length(data[[stimulus_var]])){
        if((data[i,stimulus_var]==cat1 & data[i,response_var]==cat2)|(data[i,stimulus_var]==cat2 & data[i,response_var]==cat1)){
          d[i,col]=1}
    }
  }
}
# set the column names of d to the names stored in dvars
colnames(d)=dvars
return(d)}
```

### Fit a log-linear BCM to data for stops in the CV context

```{r}
# generate d matrix
d = distmatrix(subset(rdstops, context=="CV"), stimulus_var = "stimulus", response_var = "response")
# look at d
d
# fit model
rdstopmodelCV=glm(count~stimulus+response+d,data=subset(rdstops, context=="CV"),family=poisson(link="log"))
summary(rdstopmodelCV)
```

### Recover bias parameters from the model coefficients

```{r}
# look at the coefficients
coef(rdstopmodelCV)

# extract the response coefficients - k is the base category, so the coefficient for b_k is 0
bs = c(coef(rdstopmodelCV)[4:5],0)
names(bs) = c("bp", "bt", "bk")

# exponentiate the coefficients and divide by the sum of the exponentiated coefficients
exp(bs)/sum(exp(bs))

```

## Fit BCM to stops in all contexts

The basic BCM, stimulus+response+d, can be interacted with other factors to model variation in BCM parameters across conditions

```{r}
# generate d matrix
d = distmatrix(rdstops, stimulus_var = "stimulus", response_var = "response")
# fit model
rdstopmodel=glm(count~context*(stimulus+response+d),data=rdstops,family=poisson(link="log"))
summary(rdstopmodel)
```

## Test hypotheses comparing distance coefficients

To test significance of differences between distance parameters, one strategy is to compare to a reduced model in which the relevant distance parameters are collapsed

E.g. Test the hypothesis that d_kp != d_kt, null hypothesis d_kp = d_kt

```{r}
# Manipulate d to test hypotheses about differences between distance parameters
# Starting from d matrix constructed for the stops data, rd, define d2 with a 
# dummy distance variable which has 1's for pairs p-t and p-k
d2 = d
# replace kp with the sum of kp and kt
d2[,"k.p"]=d[,"k.p"]+d[,"k.t"]
# rename kp as kpt - the name of the new dummy variable
colnames(d2)[1]="k.pt"
# delete kt, column 2
d2=d2[,-2]

# fit reduced model with d2
rdstopmodel2=glm(count~context*(stimulus+response+d2),data=rdstops,family=poisson(link="log"))

# test significance of difference using LRT
anova(rdstopmodel2,rdstopmodel,test="Chisq")
```

## Test hypotheses about bias parameters

This can be achieved via coding of the response factor.

Treatment coding (above) compares each bias parameter to the base level ("k" above)

Successive difference coding (from the MASS package) to compare two pairs of bias parameters

```{r}
library(MASS)

contrasts(rdstops$response) = contr.sdif(levels(rdstops$response))
contrasts(rdstops$response)

rdstopmodel=glm(count~context*(stimulus+response+d),data=rdstops,family=poisson(link="log"))
summary(rdstopmodel)

```

## Analyze all of the consonants in a single model

```{r}
# generate d matrix
d = distmatrix(rd, stimulus_var = "stimulus", response_var = "response")
# Fit model to all data from all contexts
rdfullmodel=glm(count~context*(stimulus+response+d), data=rd,family=poisson(link="log"))
summary(rdfullmodel)
```

Why are some distance coefficients very large but not significantly different from 0? E.g. dk.sh, dp.sh? Because there are 0's in the confusion matrix:

```{r}
filter(rd, stimulus == "k" & response == "sh" | stimulus == "sh" & response =="k")
filter(rd, stimulus == "p" & response == "sh" | stimulus == "sh" & response =="p")
filter(rd, stimulus == "t" & response == "sh" | stimulus == "sh" & response =="t")

```
