---
title: "Cumulative Logistic Regression"
format: html
editor: visual
---

```{r}
library(ordinal)
library(tidyverse)
```

# Cumulative Logistic/Probit Regression

illustrative figure

```{r}
curve(dnorm(x), xlim=c(-3,4), ylab="", xlab="")
curve(dnorm(x, mean=1), add=TRUE)
abline(v=c(-2, -1.1, 0.1, 1.5, 3),lty=2)
text(c(-2.5,-1.6,-0.5,0.9,2.2,3.5), 0.385, c("1 p","2 p","3 p","4 b","5 b","6 b"), cex = 1.3)
```

## Fit cumulative probit models to Elman's (1977) rating data on a VOT continuum

```{r}
elman = read_csv("elman.csv")

# make rating and stimulus into factors
elman$rating = factor(elman$rating)
elman$stimulus = factor(elman$stimulus)

```

Focus on a pair of stimuli that received a substantial number of different ratings: stimuli 5 and 6

```{r}
sub56 = subset(elman, stimulus=="6"|stimulus=="5")
sub56$stimulus = factor(sub56$stimulus)

# calculate qnorm(H) and qnorm(F) for each rating level, taking stimulus 5 as the 'target category'
qH = rep(NA, 5)
qF = rep(NA, 5)

for (r in 1:5){
	qH[r] = qnorm(sum(sub56[sub56$stimulus=="5" & as.numeric(sub56$rating) <= r,]$count)/sum(sub56[sub56$stimulus=="5",]$count))
	qF[r] = qnorm(sum(sub56[sub56$stimulus=="6" & as.numeric(sub56$rating) <= r,]$count)/sum(sub56[sub56$stimulus=="5",]$count))
}

#plot ROC and best fitting line
plot(qH~qF, xlab="qnorm(F)", ylab="qnorm(H)")
abline(lm(qH~qF))
```

Fit equal and unequal variance cumulative probit models to this subset of the rating data

```{r}
# equal variance model
cpm56eq = clm(rating~stimulus, weights=count, data=sub56, link="probit")
summary(cpm56eq)

# unequal variance model
cpm56uneq = clm(rating~stimulus, weights=count, data=sub56, link="probit", scale=~stimulus)
summary(cpm56uneq)
```

Extract model coefficients from unequal variances model and plot

```{r}
# rating critera
cpm56uneq$alpha
# stimulus means
cpm56uneq$beta
# ratio of standard deviations for each stimulus, relative to baseline stimulus
exp(cpm56uneq$zeta)

# plot stimulus distributions and thresholds
curve(dnorm(x,0,1),xlim=c(-2,cpm56uneq$alpha[length(cpm56uneq$alpha)]+0.5), ylim=c(0,0.7), xlab="stimuli", ylab="", xaxt="n", yaxt="n")
curve(dnorm(x,cpm56uneq$beta[],exp(cpm56uneq$zeta[])),add=TRUE)

axis(side=1, at=c(0,cpm56uneq$beta), labels =5:6)

# draw criteria
for(r in 1:length(cpm56uneq$alpha)){
  abline(v = cpm56uneq$alpha[r], lty=2)
}

# label rating regions
text((c(-2,cpm56uneq$alpha)+c(cpm56uneq$alpha,cpm56uneq$alpha[5]+1))/2, 0.6, 1:6)
```

Fit model to all of the stimuli

```{r}
# fit unequal variances cumulative probit model
cpmuneq = clm(rating~stimulus, weights=count, data=elman, link="probit", scale=~stimulus)
summary(cpmuneq)

# rating critera
cpmuneq$alpha
# stimulus means
cpmuneq$beta
# ratio of standard deviations for each stimulus, relative to baseline stimulus
exp(cpmuneq$zeta)

# compare to equal variances model
# fit equal variances cumulative probit model
cpm = clm(rating~stimulus, weights=count, data=elman, link="probit")
anova(cpmuneq, cpm)

# plot stimulus distributions and boundaries
curve(dnorm(x,0,1),xlim=c(-2,cpmuneq$beta[10]+2), ylim=c(0,0.7), xlab="stimuli", ylab="", xaxt="n", yaxt="n")
for(s in 1:length(cpmuneq$beta)){
  curve(dnorm(x,cpmuneq$beta[s],exp(cpmuneq$zeta[s])),add=TRUE)
}
axis(side=1, at=c(0,cpmuneq$beta), labels =1:11)

# draw criteria
for(r in 1:length(cpmuneq$alpha)){
  abline(v = cpmuneq$alpha[r], lty=2)
}

# label rating regions
text((c(-1,cpmuneq$alpha)+c(cpmuneq$alpha,cpmuneq$alpha[5]+1))/2, 0.6, 1:6)
```

# Ratings of imperfect rhyme acceptability

```{r}
Rhyme = read_tsv("Rhyme.txt")

# plot the relationship between rating and binned d values
rhyme_plot = ggplot(Rhyme, 
                aes(y=Response, x=as.factor(round(LuceD))))

rhyme_plot_box = rhyme_plot + geom_boxplot()
show(rhyme_plot_box)

rhyme_plot_violin = rhyme_plot + geom_violin()
show(rhyme_plot_violin)
```

Treat ratings as continuous variables: fit linear mixed effects model

```{r}
# Fit linear mixed model (singular with correlations between random effects)
# normalize predictors

library(lme4)
lmfull = lmer(Response~scale(SPETotalF)+scale(PrelimTotalF)+scale(LuceD)+
                (scale(SPETotalF)+scale(PrelimTotalF)+scale(LuceD)||Subject)+(1|Pair), data=Rhyme)
summary(lmfull)

# fit model with just d as predictor
lmLuce = lmer(Response~scale(LuceD)+
                (scale(LuceD)|Subject)+(1|Pair), data=Rhyme)
summary(lmLuce)

anova(lmfull, lmLuce)
```

Fit cumulative probit mixed effects model

```{r}
Rhyme$Response = factor(Rhyme$Response)

# singular - random slope for SPETotalF is perfectly correlated with random intercept
cpmfull = clmm(Response~scale(SPETotalF)+scale(PrelimTotalF)+scale(LuceD)+
                 (scale(SPETotalF)+scale(PrelimTotalF)+scale(LuceD)|Subject), data=Rhyme, link="probit")
summary(cpmfull)

# drop random slope for SPETotalF and add random intercept by item
cpmfull2 = clmm(Response~scale(SPETotalF)+scale(PrelimTotalF)+scale(LuceD)+
                 (scale(PrelimTotalF)+scale(LuceD)|Subject)+(1|Pair), data=Rhyme, link="probit")
summary(cpmfull2)


# fit model without LuceD
cpmred = clmm(as.factor(Response)~scale(SPETotalF)+scale(PrelimTotalF)+
                 (scale(PrelimTotalF)+scale(LuceD)|Subject)+(1|Pair), data=Rhyme, link="probit")
# compare model fits with LRT
anova(cpmfull2, cpmred)

# fit model of the relationship between rating and LuceD
cpmLuce = clmm(as.factor(Response)~scale(LuceD)+
                 (scale(LuceD)|Subject)+(1|Pair), data=Rhyme, link="probit")
summary(cpmLuce)
```
