---
title: "logistic regression"
format: html
editor: visual
---

```{r}
library(tidyverse)
```

# Binary logistic/probit regression with continuous predictor variables

## Example: Identification of a VOT continuum

```{r}
vot = read_csv("VOT.csv")
```

Fit logistic regression and probit regression models

```{r}
vot.lr = glm(cbind(ka,ga)~VOT,data=vot,family=binomial(link="logit"))
summary(vot.lr)

vot.dt = glm(cbind(ka,ga)~VOT,data=vot,family=binomial(link="probit"))
summary(vot.dt)
```

Plot the data and logistic and probit fits

```{r}
# add fitted values to data frame
#extract intercept and slope from LR model
lri=coef(vot.lr)[1]
lrs=coef(vot.lr)[2]

#extract intercept and slope from DT model
dti=coef(vot.dt)[1]
dts=coef(vot.dt)[2]

vot_plot = ggplot(vot, 
  aes(
    x = VOT,
    y = ka/100
  )) +
  geom_point() +
  geom_function(fun = \(x) 1/(1+exp(-(lrs*x+lri))),
    color = "blue") +
  geom_function(fun = \(x) pnorm(dts*x+dti),
    color = "red") +
  labs(y = "proportion of k responses", x = "VOT (ms)") +
  theme_classic(base_size = 16)
 
show(vot_plot)

```

Plot logit/probit of response probabilities.

If the models are good, these should yield straight lines

```{r}
logit_vot_plot = ggplot(vot,
  aes(x = VOT, y = log(ka/(100-ka))),
  ) +
  geom_point() +
  geom_abline(intercept = lri, slope = lrs) +
  labs(x = "VOT ms", y = "logit(p(ka))") +
  theme_classic(base_size = 16)

show(logit_vot_plot) 

probit_vot_plot= ggplot(vot,
  aes(x = VOT, y = qnorm(ka/100)),
  ) +
  geom_point() +
  geom_abline(intercept = dti, slope = dts) +
  labs(x = "VOT ms", y = "probit(p(ka))") +
  theme_classic(base_size = 16)

show(probit_vot_plot)
```

## Binary logistic regression with multiple predictor variables

Example: Identification of VOT \* f0 continuum. Data from Whalen

```{r}
# read in Whalen data
votf0 = read_csv("whalenVOTf0.csv")
```

Fit a logistic regression model with two independent variables

```{r}
votf0lrm = glm(cbind(p,b)~VOT+f0, data=votf0, family=binomial(link="logit"))
summary(votf0lrm)
```

Plot the data and fitted identification functions

```{r}
# extract model coefficients
intercept = votf0lrm$coef[1]
VOTslope = votf0lrm$coef[2]
f0slope = votf0lrm$coef[3]

# plot identification functions over data
# I couldn't work out how to plot all of the model curves in ggplot2 so this is base R

plot(p~VOT, data = votf0, col=round(f0/10), ylab = "%voiceless responses")
for (f in c(98,108,114,120,130)){
  curve(100/(1+exp(-(VOTslope*x+f0slope*f+intercept))),col=round(f/10),add=TRUE)
}

plot(log((p/100)/((1-p/100)))~VOT, data = votf0, col=round(f0/10), ylab="logit(p(voiceless))")
for (f in c(98,108,114,120,130)){
  curve(VOTslope*x+f0slope*f+intercept,col=round(f/10), add=TRUE)
}

```

## Adding inattention parameters to identification functions

Fit of the logistic regression model to the Whalen data is poor because the voiceless responses do not asymptote to 1 at high VOT (and may not asymptote to 0 at low VOT)

-   Simple logistic/probit regression models predict that identification probabilities should asymptote to 0/1 at extreme values of the linear model

-   Asymptotes other than 0 and 1 can be interpreted in terms of occasional failure by subjects to attend to the stimulus

    -   There is a probability $p_i$ that the stimulus is not attended to on a given trial

    -   When the subject does not attend to the stimulus, they guess its identity with fixed response probabilities

    -   e.g. in identification of a \[k-g\] continuum, the probability of guessing 'k' on an unattended trial is $p_k$

-    Regular detection theory model: $z(p(k)) = \beta_0 + \beta_1 VOT$

    -   $\phi (p(k))=\beta_0+\beta_1VOT$

-   With inattention parameters $p(k) = p_i p_k + (1-p_i)(\beta_0 + \beta_1 VOT)$

This model can be fit using the *psyphy* package:

-   Use function *psyfun.2asym* with link function *probit.2asym* or *logit.2asym*

```{r}
library(psyphy)

pf = psyfun.2asym(cbind(p,b)~f0+VOT,data=votf0,link=probit.2asym)

summary(pf)

int = pf$coef["(Intercept)"]
betaf0 = pf$coef["f0"]
betaVOT = pf$coef["VOT"]
lambda = pf$lambda
gamma = pf$gam
```

Note that coefficient of VOT changes substantially when inattention parameters are added

This function actually fits a model of the form

$P(x) = \gamma + ( 1 - \gamma - \lambda )f(x)$

-   where $f(x)$ is the logit or probit model

-   so $\gamma$ is the lower asymptote and $1-\lambda$ is the upper asymptote

    -   $\gamma = p_i p_k$ and $\gamma + \lambda = p_i$

```{r}
int = pf$coef["(Intercept)"]
betaf0 = pf$coef["f0"]
betaVOT = pf$coef["VOT"]
lambda = pf$lambda
gamma = pf$gam

# p_i
gamma + lambda
# p_k
gamma/(gamma + lambda)

plot(p/100~VOT, data = votf0, col=round(f0/10))
for (f in c(98,108,114,120,130)){
  curve((1-lambda-gamma)*pnorm(int+betaf0*f+betaVOT*x)+gamma,col=round(f/10), add=TRUE)
}
```

# Multinomial logistic regression

Example: Identification of a VOT continuum with three categories b, p, ph, by Thai listeners (Lisker & Abramson 1970)

```{r}
# Read in the Lisker & Abramson Thai data
thai = read_csv("thai.csv")

# plot the data
thai_plot = ggplot(thai,
  aes(
    x = VOT,
    y = p/(b+p+ph)
  )) +
  geom_point(position = "jitter") +
  geom_point(aes(
    x = VOT,
    y = b/(b+p+ph),
  ), color = "red") +
  geom_point(aes(
    x = VOT,
    y = ph/(b+p+ph)
  ), color = "blue") +
  labs(x = "VOT (ms)", y = "response proportion") +
  theme_classic(base_size = 16)

show(thai_plot)

```

## vglm

fit multinomial logistic regression model to the Thai data using vglm from the VGAM package

```{r}
library(VGAM)

# multinomial logistic regression with VGAM
# base category is the last
thai.vglm = vglm(cbind(b, ph, p) ~ VOT, data=thai, family = multinomial())
summary(thai.vglm)
```

Add model fitted curves to the plot

```{r}
#extract intercepts and slopes from MLR model
intb=coef(thai.vglm)[1]
slopeb=coef(thai.vglm)[3]

intph=coef(thai.vglm)[2]
slopeph=coef(thai.vglm)[4]

thai_plot = thai_plot +
  geom_function(fun = \(x) exp(intb+slopeb*x)/(exp(intb+slopeb*x)+1+exp(intph+slopeph*x)), color = "red") +
  geom_function(fun = \(x) exp(intph+slopeph*x)/(exp(intb+slopeb*x)+1+exp(intph+slopeph*x)), col = "blue") +
  geom_function(fun = \(x) 1/(exp(intb+slopeb*x)+1+exp(intph+slopeph*x)))

show(thai_plot)
```

```{r}
# plot log odds and logit models
thai_logit_plot = ggplot(thai,
  aes(
    x = VOT,
    y = log(b/p)
  )) +
  geom_point(color = "red") +
  geom_point(aes(
    x = VOT,
    y = log(p/p)
  )) +
  geom_point(aes(
    x = VOT,
    y = log(ph/p)
  ), color = "blue") +
  theme_classic(base_size = 16)

show(thai_logit_plot)

# add model lines
thai_logit_plot = thai_logit_plot +
  geom_abline(intercept = intb, slope = slopeb, color = "red") +
  geom_abline(intercept = intph, slope = slopeph, color = "blue") +
  geom_hline(yintercept = 0)

show(thai_logit_plot)



```

## nnet

fit multinomial logistic regression model to the Thai data using *multinom* from the *nnet* package

```{r}
library(nnet)

# base category is the first
thai.mn = multinom(cbind(p, b, ph)~VOT, data=thai)
summary(thai.mn)

# compare to vglm
summary(thai.vglm)
```

## Reformulate multinomial logistic regression model as a log-linear model

```{r}
# reshape thai data to longer form with counts of each stimulus-response combination
thai_long = pivot_longer(thai, cols = c("b", "p", "ph"), names_to = "response", values_to = "count")
# add stimulus factor with one level of 
thai_long = mutate(thai_long, stimulus = factor(thai_long$VOT))
# make p the baseline category for the reponse variable, to match the logistic regression analyses above
thai_long$response = factor(thai_long$response)
contrasts(thai_long$response) = contr.treatment(levels(thai_long$response), base = 2)

thai.ll = glm(count~stimulus+response*VOT, data=thai_long, family=poisson(link="log"))
summary(thai.ll)

```

-   The intercepts of the multinomial logistic regression model correspond to the coefficients of the response variable

-   The coefficients of VOT in multinomial logistic regression models correspond to the coefficients of responseC:VOT interactions in the loglinear model

    -   coefficient for the baseline category p is fixed at 0 (hence NA in the model summary)

-   In general, any variables that affect response probabilities must be entered into the loglinear model as interactions with the response variable

-   Ignore stimulus coefficients
